***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**

### 2015-10-08 Schloss_Cluster_PeerJ_2015.Rmd
Write abstract

## To Do:
* references
* abstract
* revise

@@

## Abstract

>>>>

## To Do:
* references
* revise

@@

\newpage
## Abstract

**Background.** 16S rRNA gene sequences are routinely assigned to operational taxonomic units (OTUs) that are then used to analyze complex microbial communities. A number of methods have been employed to carry out the assignment of 16S rRNA gene sequences to OTUs leading to confusion over which method is the best. A recent study has suggested that a clustering method should be selected based on its ability to generate stable OTU assignments as additional sequences are added to the dataset rather than the ability of the method to properly represent the distances between the sequences. It was thus necessary to replicate the previous analysis and assess the quality of the OTU assignments.

**Methods.** Our analysis implemented five *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering and two reference-based methods including the open and closed-reference methods. By varying the number of sequences sampled from each of two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the stability of OTU assignments and their quality.

**Results.** Our analysis demonstrated that the stability of OTU assignments does not predict to the ability of the methods to reflect the distance between the sequences. We found that the average linkage and the distance and abundance-based greedy clustering methods out performed the open and closed-reference methods. Depending on the dataset being analyzed, the average linkage or greedy clustering algorithms provided the best dataset. Further exploration of the reference-based methods indicated that when using USEARCH to identify the closest reference the results were sensitive to the order of the reference sequences. When USEARCH was replaced by VSEARCH the quality and stability of the OTU assignments by the reference-based methods were improved. Finally, we demonstrated that for the greedy algorithms VSEARCH produced assignments that were comparable to those produced by USEARCH making VSEARCH a viable free and open source alternative to USEARCH.

**Discussion.** This study demonstrates that the quality of clustering assignments needs to be assessed for multiple methods per dataset to identify the optimal clustering method for the dataset. The analysis also calls into question the quality and stability of OTU assignments generated by the open and closed-reference methods as implemented in current versions of QIIME.
